Parsimonious Representation of Nonlinear Dynamical Systems Through Manifold Learning: A Chemotaxis Case Study

### 2015-05-22 chemotaxis/paper/manuscript.tex
updated chemotaxis paper for submission

To review and illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}.
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
respectively, on the domain, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide an estimate of the
relative magnitude of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases,
the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases
(this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper.
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
and a proper parameterization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
smallest $d$ eigenvalues.
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any
additional directions intrinsic to the strip geometry; we will refer to these as ``repeated eigendirections,''
and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
Clearly, this problem of repeated eigendirections also arises for more curved of more than two variables.
%
We note that, although the eigenfunctions can only be written analytically for very special cases,
it has been observed empirically that they often provide appropriate coordinates
to usefully parametrize more complicated, nonlinear manifolds.
%

@@

In most applications, we are not given a description of the continuous manifold on which the data lie.
%
Instead, we are given data {\em sampled} from the underlying manifold,
and the parameterization of the manifold needs to be uncovered from the data.
%
This can be accomplished by constructing a matrix which approximates the Laplace-Beltrami operator.
%
It was shown in \cite{coifman2006geometric} that, in the limit of infinite data,
this discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold.
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator.

@@

As discussed previously, the eigenfunctions provide a parametrization of the manifold,
such that $\mu_j^\tau \phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$, where $\tau \ge 0$ is
a parameter (in our examples, we will take $\tau=0$).
%
This is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}, where
we order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
%
Because the matrix $\mathbf{A}$ is row-stochastic ($\sum_j \mathbf{A}_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
%
The distance induced by this mapping is called the standard diffusion distance,
%
\begin{equation}
D^2_\tau(z(i), z(j)) = \sum_{k=1}^{m-1} \mu_k^{2 \tau} \left( \phi_k(i) - \phi_k(j)  \right)^2.
\end{equation}
%
Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows
data sampled from a strip, colored by eigenvectors of $\mathbf{A}$.
%
In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$,
and thus parametrize the manifold.
%
%TODO: maybe we can put here the new figures for the different values of $\alpha$ instead of these figures.
%
Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{1.5in}
\centering
\includegraphics[height=0.75in]{torus1}
\includegraphics[height=1in]{torus1_evals}
\caption{}
\end{subfigure}
%
%\hfill
%
\begin{subfigure}{1.5in}
\centering
\includegraphics[height=0.75in]{torus2}
\includegraphics[height=1in]{torus2_evals}
\caption{}
\end{subfigure}
%
%\hfill
%
\begin{subfigure}{1.5in}
\centering
\includegraphics[height=0.75in]{torus3}
\includegraphics[height=1in]{torus3_evals}
\caption{}
\end{subfigure}
%
\hfill
%
\caption{Tori data sets (top, see \eqref{eq:torus}) and corresponding diffusion maps eigenvalues (bottom) for (a) $r_1 = 3$, $r_2 = 1$. (b) $r_1 = 5$, $r_2 = 1$. (c) $r_1 = 10$, $r_2 = 1$. The tori are parameterized by two angles: $\theta_1$, which parameterizes the large outer circle, and $\theta_2$, which parameterizes the smaller inner circle. In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$. The second pair of unique eigendirections (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
%
\label{fig:torus}
%
\end{figure}

@@

The microscopic model consists of a collection of $N$ cells (for our simulations, we take $N=1000$) whose states are 
defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of cell $i$ at time $t$.
%
The velocity of each cell is $\pm s$, where $s$ is a (fixed) speed.
%
We initialize the cells such that
\begin{equation}\label{eqn:system}
\begin{aligned}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{aligned}
\end{equation}
where $0 < p < 1$ is the probability of a cell initially moving to the right.
%
The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$ (this switching is physically controlled by extracellular signals).
%
For our specific simulations, we set $s^2/\lambda=1$, which is consistent with the analysis presented in \cite{othmer1988models}.
%
We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate our main points;
however, our analysis is not restricted to such specific cases and more complex sets of initial conditions could be used.
%
Each data set consists of $10$ stochastic simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
%
We allow each simulation to evolve for $t_{max}$ time units, and record the positions of the $N$ cells at $\Delta t$ intervals; we use data with $t > 0$ for analysis.
%
We will show that the parameter $t_{obs} \equiv t_{max}/N$, which determines the time scale of observation, is of critical importance when we address the limiting asymptotic behavior.

@@

For the chemotaxis example, we use histograms of the cell positions as observers.
%
Histograms are invariant to the indexing of the cells, while retaining information about the relative spatial locations of the cells.
%
Histograms are also robust to noise \cite{talmon2013empirical}.
%
Instead of the standard Euclidean distance, we use the Earth Mover's Distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms.
%
Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%
It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are from each other.
%
Although the brute-force computation of the EMD is computationally expensive, 
there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009, leeb2014lipschitz}.
%
For the specific case of spatially one-dimensional data, the EMD is equivalent to the $L_1$-norm between the 
cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
\begin{equation}
\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|;
\end{equation}
the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin.
%
Here, we choose to only use the positions of the cells in the distance computation; utilizing both the positions and velocities produces qualitatively similar results (not shown).
%
We note that although histograms are not essential to the theory of EMD, they make the computation practical.

@@

Guided by the known theory of the model macroscopic PDE system, we consider two asymptotic regimes for the simulation.
%
When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
\begin{equation}
\begin{aligned}
\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
\end{aligned}
\end{equation}
%
When $\lambda \rightarrow \infty$ with $s^2/\lambda$ fixed, \eqref{eq:second_order_pde} approaches the diffusion equation,
\begin{equation} \label{eq:diff_eqn}
2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2},
\end{equation}
%
where $D=s^2/\lambda$.
%
The above analysis shows that the initial distribution of velocities of the cells 
(determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
%
When $\lambda \rightarrow 0$, the dynamics are described by the telegrapher equation, and the effect of the initial velocity distribution 
persists throughout the trajectory.
%
When $\lambda \rightarrow \infty$, the dynamics are described by a single heat equation, and the initial conditions 
for the velocity are not especially important, since the velocity distribution quickly equilibrates and we see purely diffusive behavior.

The relative importance of $p$ and $t$ from the analytic description are consistent with the results 
in Figure~\ref{fig:dmaps_embed_emd}, where $p$ is correlated with the dominant diffusion map coordinate 
when $\lambda$ is small, and becomes correlated with the subdominant coordinate when $\lambda$ is large.
%
Furthermore, in the small $\lambda$ regime (wave equation), shown in 
Figures \ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}, the points corresponding to 
small times are more tightly clustered than the points corresponding to large times.
%
This is in agreement with the macroscopic model: at small times, the cells are more 
condensed around $x=0$, and it is more difficult to distinguish the cells moving to the left from the cells moving to the right.
%%%YGK
%%% that is because we did not use the velocities in our diffusion maps.....
%%%
%
On the other hand, at large times, once the cells move away from the origin, this separation between left-moving and
right-moving cells is clear.
%
For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}, 
we observe that at small times (before the cell velocities have sufficient opportunity to switch and equilibrate), the initial velocity distribution $p$ can be well perceived in the
embedding in Figure \ref{subfig:large_lambda_p}, as the initial imbalance in left-right velocities affects
the initial displacements.
%
On the other hand, for longer times, we observe that the initial distribution $p$ cannot be well perceived
in the embedding in Figure \ref{subfig:large_lambda_p}, 
as the velocities have equilibrated and their initial distribution is practically forgotten.
%
Overall, in both cases, we obtain, in an unsupervised data-driven manner, a useful low-dimensional embedding of the data.
%
Correlating this embedding with potential candidate macroscopic observables (such as time, densities, fluxes, and initial condition statistics) helps elucidate
an accurate picture of the macroscopic variables that govern the system dynamics.

@@

\begin{figure}[t]
%
\centering
\includegraphics[width=0.5\textwidth]{tmax_lambda_transition}
%
\caption{Detecting the change in dimensionality. The appropriate ratio the two eigenvalues (see \eqref{eq:chemotaxis_eval_ratio}) which correspond to unique eigendirections is plotted as a function of $t_{max}$ and $\lambda$. Each data point is the average eigenvalue ratio over three replicate data sets, and in each simulation, $\Delta t=t_{max}/10$. The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions. The curve $1/\lambda = t_{max}/N$ is shown in white. }
%
\label{fig:chemotaxis_compare_timescales_evals}
%
\end{figure}

@@

Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as 
a function of $\lambda$ and the time scale of observation $t_{max}$.
%
From this eigenvalue ratio, we can easily detect changes in the estimated dimensionality as we vary the relevant parameters.
%
For small $\lambda$ and/or short $t_{obs} = t_{max} / N$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics.
%
However, when $\lambda$ and/or $t_{obs}$ is large, the system rapidly equilibrates and the data are effectively one-dimensional, parametrized by time.
%
For this specific example, these changes are consistent with the analytically-predicted  shift in the dynamical behavior from the telegrapher equation to the diffusion equation.
%
From \eqref{eq:second_order_pde}, we can see that $t_{obs}$, the time scale of observation, should be larger than the time scale $1/\lambda$ for the diffusion equation limit in \eqref{eq:diff_eqn} to hold; we therefore expect a transition in the dynamical behavior at
\begin{equation}
t_{obs} \approx \frac{1}{\lambda}.
\end{equation}
%
This curve is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} and is consistent with the transition predicted by the diffusion maps eigenvalues.

>>>>

To review and illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
%
The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
\begin{equation} \label{eq:evals}
\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
\end{equation}
for $k_1, k_2 = 0, 1, 2, \dots$,
and the corresponding eigenfunctions are
\begin{equation} \label{eq:efuncs}
\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
\end{equation}
where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}.
%
We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and
$\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
respectively, on the domain, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
%
Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide an estimate of the
relative magnitude of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases,
the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases
(this will be discussed further in Section~\ref{sec:relative_lengths}).
%
The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper.
%
First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
and a proper parametrization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
smallest $d$ eigenvalues.
% TODO: address this in later section
%
Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any
additional directions intrinsic to the strip geometry; we will refer to these as ``repeated eigendirections,''
and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
%
Clearly, this problem of repeated eigendirections also arises for more curved of more than two variables.
%
We note that, although the eigenfunctions can only be written analytically for very special cases,
it has been observed empirically that they often provide appropriate coordinates
to usefully parametrize more complicated, nonlinear manifolds.
%

@@

In most applications, we are not given a description of the continuous manifold on which the data lie.
%
Instead, we are given data {\em sampled} from the underlying manifold,
and the parametrization of the manifold needs to be uncovered from the data.
%
This can be accomplished by constructing a matrix which approximates the Laplace-Beltrami operator.
%
It was shown in \cite{coifman2006geometric} that, in the limit of infinite data,
this discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold.
%
As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator.

@@

As discussed previously, the eigenfunctions provide a parametrization of the manifold,
such that $\mu_j^\tau \phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$, where $\tau \in \mathbb{N}$ is
a parameter (in our examples, we will take $\tau=0$).
%
The parameter $\tau$ corresponds to the number of iterates of the Laplace operator, or in other words, the number of diffusion steps (for more details, see \cite{coifman2006geometric}). 
%
%In practice, it has a similar, but not equivalent, effect as the kernel scale \epsilon. 
% 
%
This mapping $z(i) \mapsto \left(\mu_1^\tau \phi_{1}(i), \dots, \mu_{m-1}^\tau \phi_{m-1}(i)\right)$ is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}, where
we order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
%
Because the matrix $\mathbf{A}$ is row-stochastic ($\sum_j \mathbf{A}_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
%
The distance induced by this mapping is called the standard diffusion distance,
%
\begin{equation}
D^2_\tau(z(i), z(j)) = \sum_{k=1}^{m-1} \mu_k^{2 \tau} \left( \phi_k(i) - \phi_k(j)  \right)^2.
\end{equation}
%
Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows
data sampled from a strip, colored by eigenvectors of $\mathbf{A}$.
%
In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$,
and thus parametrize the manifold.
%
%TODO: maybe we can put here the new figures for the different values of $\alpha$ instead of these figures.
%
Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.

@@

\begin{figure}[t]
\centering
\begin{subfigure}{1.5in}
\centering
\includegraphics[height=0.75in]{torus1}
\includegraphics[height=1in]{torus1_evals}
\caption{}
\end{subfigure}
%
%\hfill
%
\begin{subfigure}{1.5in}
\centering
\includegraphics[height=0.75in]{torus2}
\includegraphics[height=1in]{torus2_evals}
\caption{}
\end{subfigure}
%
%\hfill
%
\begin{subfigure}{1.5in}
\centering
\includegraphics[height=0.75in]{torus3}
\includegraphics[height=1in]{torus3_evals}
\caption{}
\end{subfigure}
%
\hfill
%
\caption{Tori data sets (top, see \eqref{eq:torus}) and corresponding diffusion maps eigenvalues (bottom) for (a) $r_1 = 3$, $r_2 = 1$. (b) $r_1 = 5$, $r_2 = 1$. (c) $r_1 = 10$, $r_2 = 1$. The tori are parametrized by two angles: $\theta_1$, which parametrizes the large outer circle, and $\theta_2$, which parametrizes the smaller inner circle. In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$. The second pair of unique eigendirections (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
%
\label{fig:torus}
%
\end{figure}

@@

The microscopic model consists of a collection of $N$ cells (for our simulations, we take $N=1000$) whose states are 
defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of cell $i$ at time $t$.
%
The velocity of each cell is $\pm s$, where $s$ is a (fixed) speed.
%
We initialize the cells such that
\begin{equation}\label{eqn:system}
\begin{aligned}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{aligned}
\end{equation}
where $0 < p < 1$ is the probability of a cell initially moving to the right.
%
The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$ (this switching is physically controlled by extracellular signals).
%
For our specific simulations, we set $s^2/\lambda=1$, which is consistent with the analysis presented in \cite{othmer1988models}.
%
We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate our main points;
however, our analysis is not restricted to such specific cases and more complex sets of initial conditions could be used.
%
Each data set consists of $10$ stochastic simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
%
We allow each simulation to evolve for $t_{max}$ time units, and record the states of the $N$ cells every $\Delta t$ time units; we use data with $t > 0$ for analysis.
%
We will show that the parameter $t_{obs} \equiv t_{max}/N$, which determines the time scale of observation, is of critical importance when we address the limiting asymptotic behavior.

@@

For the chemotaxis example, we use histograms of the cell positions as observers.
%
Histograms are invariant to the indexing of the cells, while retaining information about the relative spatial locations of the cells.
%
Histograms are also robust to noise \cite{talmon2013empirical}.
%
Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms.
%
Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
%
It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are from each other.
%
Although the brute-force computation of the EMD is computationally expensive, 
there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009, leeb2014lipschitz}.
%
For the specific case of spatially one-dimensional data, the EMD is equivalent to the $L_1$-norm between the 
cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
\begin{equation}
\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|;
\end{equation}
the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin.
%
Here, we choose to only use the positions of the cells in the distance computation; utilizing both the positions and velocities produces qualitatively similar results (not shown).
%
We note that although histograms are not essential to the theory of EMD, they make the computation practical.

@@

Guided by the known theory of the model macroscopic PDE system, we consider two asymptotic regimes for the simulation.
%
When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
\begin{equation}
\begin{aligned}
\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
\end{aligned}
\end{equation}
%
When $\lambda \rightarrow \infty$ with $s^2/\lambda$ fixed, \eqref{eq:second_order_pde} approaches the diffusion equation,
\begin{equation} \label{eq:diff_eqn}
2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2},
\end{equation}
%
where $D=s^2/\lambda$.
%
The above analysis shows that the initial distribution of velocities of the cells 
(determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
%
When $\lambda \rightarrow 0$, the dynamics are described by the telegrapher equation, and the effect of the initial velocity distribution 
persists throughout the trajectory.
%
When $\lambda \rightarrow \infty$, the dynamics are described by a single diffusion equation, and the initial conditions 
for the velocity are not especially important, since the velocity distribution quickly equilibrates and we see purely diffusive behavior.

The relative importance of $p$ and $t$ from the analytic description are consistent with the results 
in Figure~\ref{fig:dmaps_embed_emd}, where $p$ is correlated with the dominant diffusion map coordinate 
when $\lambda$ is small, and becomes correlated with the subdominant coordinate when $\lambda$ is large.
%
Furthermore, in the small $\lambda$ regime (wave equation), shown in 
Figures \ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}, the points corresponding to 
small times are more tightly clustered than the points corresponding to large times.
%
This is in agreement with the macroscopic model: at small times, the cells are more 
condensed around $x=0$, and it is more difficult to distinguish the cells moving to the left from the cells moving to the right.
%%%YGK
%%% that is because we did not use the velocities in our diffusion maps.....
%%%
%
On the other hand, at large times, once the cells move away from the origin, this separation between left-moving and
right-moving cells is clear.
%
For the large $\lambda$ case (diffusion equation), shown in Figures \ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}, 
we observe that at small times (before the cell velocities have sufficient opportunity to switch and equilibrate), the initial velocity distribution $p$ can be well perceived in the
embedding in Figure \ref{subfig:large_lambda_p}, as the initial imbalance in left-right velocities affects
the initial displacements.
%
On the other hand, for longer times, we observe that the initial distribution $p$ cannot be well perceived
in the embedding in Figure \ref{subfig:large_lambda_p}, 
as the velocities have equilibrated and their initial distribution is practically forgotten.
%
Overall, in both cases, we obtain, in an unsupervised data-driven manner, a useful low-dimensional embedding of the data.
%
Correlating this embedding with potential candidate macroscopic observables (such as time, densities, fluxes, and initial condition statistics) helps elucidate
an accurate picture of the macroscopic variables that govern the system dynamics.

@@

\begin{figure}[t]
%
\centering
\includegraphics[width=0.5\textwidth]{tmax_lambda_transition}
%
\caption{Detecting the change in dimensionality. The appropriate ratio the two eigenvalues (see \eqref{eq:chemotaxis_eval_ratio}) which correspond to unique eigendirections is plotted as a function of $t_{obs}$ and $\lambda$. Each data point is the average eigenvalue ratio over three replicate data sets, and in each simulation, $\Delta t=t_{max}/10$. The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions. The curve $t_{obs} = 1/\lambda$ is shown in white. }
%
\label{fig:chemotaxis_compare_timescales_evals}
%
\end{figure}

@@

Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as 
a function of $\lambda$ and the time scale of observation $t_{obs} = t_{max} / N$.
%
From this eigenvalue ratio, we can easily detect changes in the estimated dimensionality as we vary the relevant parameters.
%
For small $\lambda$ and/or short $t_{obs}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics.
%
However, when $\lambda$ and/or $t_{obs}$ is large, the system rapidly equilibrates and the data are effectively one-dimensional, parametrized by time.
%
For this specific example, these changes are consistent with the analytically-predicted  shift in the dynamical behavior from the telegrapher equation to the diffusion equation.
%
From \eqref{eq:second_order_pde}, we can see that $t_{obs}$, the time scale of observation, should be larger than the time scale $1/\lambda$ for the diffusion equation limit in \eqref{eq:diff_eqn} to hold; we therefore expect a transition in the dynamical behavior at
\begin{equation}
t_{obs} \approx \frac{1}{\lambda}.
\end{equation}
%
This curve is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} and is consistent with the transition predicted by the diffusion maps eigenvalues.
